Post-replication citation patterns in psychology: Four case studies

### 2021-02-07 manuscript/manuscript.Rmd
applied styler to analysis files

abstract: >
   Replication studies that contradict published findings may facilitate scientific self-correction by triggering a reappraisal of the original findings; however, the research community's response to replication studies has not been studied systematically. One approach for gauging responses to replication results is to examine how they impact citations to original studies. In this study, we explored post-replication citation patterns in the context of four prominent multi-laboratory replication attempts published in the field of psychology that strongly contradicted and outweighed prior findings. Generally, we observed a small post-replication decline in the number of favourable citations and a small increase in unfavourable citations. This indicates only modest corrective effects and implies considerable perpetuation of belief in the original findings. Replication results that strongly contradict an original finding do not necessarily nullify its credibility; however, one might at least expect the replication results to be acknowledged and explicitly debated in subsequent literature. By contrast, we found substantial citation bias: the majority of articles citing the original studies neglected to cite the relevant replication. Of those articles that did cite the replication, but continued to cite the original study favourably, approximately half offered an explicit defence of the original study. Our findings suggest that even replication results that strongly contradict original findings do not necessarily prompt a corrective response from the research community.

@@

Table \@ref(tab:patterns) outlines several citation patterns that might follow a contradictory replication result, each reflecting different types of response by the research community. We have tentatively categorised these response patterns as being ‘progressive’ or ‘regressive’ depending on their expected impact on the accumulation of scientific knowledge[^footnote2]. The first set of patterns, belief correction/perpetuation, refer to what is often considered the primary functional role of (contradictory) replication studies - to change belief in the credibility of exaggerated or erroneous original findings (Ioannidis, 2012; Vazire & Holcombe, 2020; Zwaan et al., 2018). In the absence of a explicit defence of an original finding that convincingly explains a strongly contradictory result that outweighs an original finding (see ‘explicit defence’ below), a progressive response might involve a decrease in favourable citations and an increase in unfavourable citations, reflecting updated beliefs about the credibility of the original finding. Conversely, a regressive response might involve maintenance of (or even increase in) favourable citations, and relatively few unfavourable citations, suggesting a perpetuation of belief in the credibility of the original finding despite the contradictory replication result. Prior research has documented how favourable citations to observational epidemiology studies can persist despite the claims of those studies being strongly contradicted in subsequent randomized trials (Tatsioni et al., 2007). Similarly, it has been reported that even when articles are retracted, they can continue to receive favourable citations (Budd et al., 1998). Thus, there is evidence that belief in the credibility of original findings can perpetuate even when subsequent events cast doubt on their credibility; however, we are unaware of similar evidence in the context of replication studies that were explicitly designed to test the reliability of prior findings.

@@

```{r patterns}
tibble(
  "Progressive responses" = c("*Belief correction*
A decrease in the number of favourable citations may reflect a decline in belief in the credibility of the original finding (a belief correction pattern). This may be accompanied by a relative increase in the number of unfavourable citations (an active belief correction pattern) or relatively fewer/no unfavourable citations (a passive belief correction pattern).", "*Citation balance*
Articles that cite the original study may also cite the contradictory replication, reflecting that relevant evidence has been considered (a balanced citation pattern).", "*Explicit defence*
Articles that favourably cite the original study and unfavourably cite the contradictory replication may offer concrete counter-arguments that state why the credibility of the original finding has not been undermined (an explicit defence pattern)."),
  "Regressive responses" = c("*Belief perpetuation*
A maintenance of, or increase in, the number of favourable citations may reflect that a maintenance of or increase in belief in the credibility of the original finding (a belief perpetuation pattern). This may be accompanied by a relative increase in unfavourable citations (a challenged belief perpetuation pattern) or relatively fewer/no unfavourable citations (an unchallenged belief perpetuation pattern).", "*Citation bias*
Articles that cite the original study may neglect to also cite the contradictory replication, reflecting that relevant evidence has been neglected, either through lack of awareness or deliberate omission (a citation bias pattern).", "*Absent defence*
Articles that favourably cite the original study and unfavourably cite the contradictory replication may offer no concrete counter-arguments that state why the credibility of the original finding has not been undermined (an absent defence pattern).")
) %>%
  kable(booktabs=T,longtable=T,caption = "Progressive or regressive responses to strongly contradictory replication results and their expected impact on citation patterns for original studies.") %>%
  column_spec(1,'8.5cm') %>%
  column_spec(2,'8.5cm')
```

@@

# author surnames for counter-argument papers
ca_auth_list <- d_args %>%
  distinct(doi, .keep_all = T) %>%
  pull(authors) %>%
  str_c(collapse = '; ')

# frequency of authors in counter-arg papers
ca_auth_freq <- ca_auth_list %>%
  str_split(';') %>%
  table(dnn = 'name') %>%
  as.data.frame() %>%
  arrange(desc(Freq)) %>%
  rowwise() %>%
  mutate(surname = str_split(name,',')[[1]][1]) # extract surnames

# frequency of original authors also authoring counter-arg papers
ca_auth_surnames <- ca_auth_freq$surname %>% str_c(collapse = ' ')
org_auths_ca <- org_auths_list[str_detect(ca_auth_surnames, org_auths_list)]

@@

# Materials, data, and analysis code availability
All data, materials, and analysis scripts related to this study are publicly available on the Open Science Framework (https://osf.io/w8h2q/). To facilitate reproducibility this manuscript was written by interleaving regular prose and analysis code using knitr (Xie, 2017) and papaja (Aust & Barth, 2020), and is available in a Code Ocean container (TO BE ADDED) which re-creates the software environment in which the original analyses were performed.

>>>>

abstract: >
   Replication studies that contradict prior findings may facilitate scientific self-correction by triggering a reappraisal of the original studies; however, the research community's response to replication results has not been studied systematically. One approach for gauging responses to replication results is to examine how they impact citations to original studies. In this study, we explored post-replication citation patterns in the context of four prominent multi-laboratory replication attempts published in the field of psychology that strongly contradicted and outweighed prior findings. Generally, we observed a small post-replication decline in the number of favourable citations and a small increase in unfavourable citations. This indicates only modest corrective effects and implies considerable perpetuation of belief in the original findings. Replication results that strongly contradict an original finding do not necessarily nullify its credibility; however, one might at least expect the replication results to be acknowledged and explicitly debated in subsequent literature. By contrast, we found substantial citation bias: the majority of articles citing the original studies neglected to cite relevant replication results. Of those articles that did cite the replication, but continued to cite the original study favourably, approximately half offered an explicit defence of the original study. Our findings suggest that even replication results that strongly contradict original findings do not necessarily prompt a corrective response from the research community.

@@

Table \@ref(tab:patterns) outlines several citation patterns that might follow a contradictory replication result, each reflecting different types of response by the research community. We have tentatively categorised these response patterns as being ‘progressive’ or ‘regressive’ depending on their expected impact on the accumulation of scientific knowledge[^footnote2]. The first set of patterns, belief correction/perpetuation, refer to what is often considered a primary functional role of (contradictory) replication studies - to change belief in the credibility of exaggerated or erroneous original findings (Ioannidis, 2012; Vazire & Holcombe, 2020; Zwaan et al., 2018). In the absence of an explicit defence of an original study that convincingly explains a strongly contradictory replication result (see ‘explicit defence’ below), a progressive response might involve a decrease in favourable citations and an increase in unfavourable citations, reflecting updated beliefs about the credibility of the original finding. Conversely, a regressive response might involve maintenance of (or even increase in) favourable citations, and relatively few unfavourable citations, suggesting a perpetuation of belief in the credibility of the original finding despite the contradictory replication result. Prior research has documented how favourable citations to observational epidemiology studies can persist despite the claims of those studies being strongly contradicted in subsequent randomized trials (Tatsioni et al., 2007). Similarly, it has been reported that even when articles are retracted, they can continue to receive favourable citations (Budd et al., 1998). Thus, there is evidence that belief in the credibility of original findings can perpetuate even when subsequent events cast doubt on their credibility; however, we are unaware of similar evidence in the context of replication studies that were explicitly designed to test the reliability of prior findings.

@@

```{r patterns}
tibble(
  "Progressive responses" = c("*Belief correction*
A decrease in the number of favourable citations may reflect a decline in belief in the credibility of the original finding (a belief correction pattern). This may be accompanied by a relative increase in the number of unfavourable citations (an active belief correction pattern) or relatively fewer/no unfavourable citations (a passive belief correction pattern).", "*Citation balance*
Articles that cite the original study may also cite the contradictory replication, reflecting that relevant evidence has been considered (a balanced citation pattern).", "*Explicit defence*
Articles that favourably cite the original study and unfavourably cite the contradictory replication may offer concrete counter-arguments that state why the credibility of the original finding has not been undermined (an explicit defence pattern)."),
  "Regressive responses" = c("*Belief perpetuation*
A maintenance of, or increase in, the number of favourable citations may reflect that a maintenance of or increase in belief in the credibility of the original finding (a belief perpetuation pattern). This may be accompanied by a relative increase in unfavourable citations (a challenged belief perpetuation pattern) or relatively fewer/no unfavourable citations (an unchallenged belief perpetuation pattern).", "*Citation bias*
Articles that cite the original study may neglect to also cite the contradictory replication, reflecting that relevant evidence has been neglected, either through lack of awareness or deliberate omission (a citation bias pattern).", "*Absent defence*
Articles that favourably cite the original study and unfavourably cite the contradictory replication may offer no concrete counter-arguments that state why the credibility of the original finding has not been undermined (an absent defence pattern).")
) %>%
  kable(booktabs = T, longtable = T, caption = "Progressive or regressive responses to strongly contradictory replication results and their expected impact on citation patterns for original studies.") %>%
  column_spec(1, "8.5cm") %>%
  column_spec(2, "8.5cm")
```

@@

# author surnames for counter-argument papers
ca_auth_list <- d_args %>%
  distinct(doi, .keep_all = T) %>%
  pull(authors) %>%
  str_c(collapse = "; ")

# frequency of authors in counter-arg papers
ca_auth_freq <- ca_auth_list %>%
  str_split(";") %>%
  table(dnn = "name") %>%
  as.data.frame() %>%
  arrange(desc(Freq)) %>%
  rowwise() %>%
  mutate(surname = str_split(name, ",")[[1]][1]) # extract surnames

# frequency of original authors also authoring counter-arg papers
ca_auth_surnames <- ca_auth_freq$surname %>% str_c(collapse = " ")
org_auths_ca <- org_auths_list[str_detect(ca_auth_surnames, org_auths_list)]

@@

# Materials, data, and analysis code availability
All data, materials, and analysis scripts related to this study are publicly available on the Open Science Framework (https://osf.io/w8h2q/). To facilitate reproducibility this manuscript was written by interleaving regular prose and analysis code using knitr (Xie, 2017) and papaja (Aust & Barth, 2020), and is available in a Code Ocean container (https://doi.org/10.24433/CO.4225975.v1) which re-creates the software environment in which the original analyses were performed.
